Human capital mediates natural selection in contemporary humans

### 2022-01-03 why-natural-selection.Rmd
Tweaks to text

We created polygenic scores for 33 traits in `r comma_num(nrow(famhist))`
individuals of European descent, corrected for ancestry using 100 genetic
principal components (see Materials and Methods). Figure
\@ref(fig:plot-means-over-time) plots mean polygenic scores in the sample by
5-year birth intervals. Several scores show consistent increases or declines
over this 30-year period, of the order of 5% of a standard deviation. These
changes could reflect natural selection within the UK population, but also
emigration, or ascertainment bias within the sample. Respondents have higher
income and are better educated than the UK population, and they may also differ
on other unobserved characteristics [@10.1093/aje/kwx246]. Since richer and more
educated people also live longer, this bias could also increase with age.
Lastly, the sample sex ratio skews `r female_ratio * 100`% female.

@@

To show this, we write down a simple model of fertility choices. $h$ is
an individual's level of human capital. For now, we simply identify this with
his or her wage $W$. Raising a child takes time $b$. People maximize utility
from the number of children $N$ and from income $Y\equiv(1-bN)W$. An individual's
payoff is

@@

@@

@@

Second, we directly test whether the correlation between each polygenic score
and fertility is mediated by educational attainment (Appendix Table
\@ref(tab:tbl-res-mediation)). For `r n_pos_sig` scores, the indirect effect of
the score on fertility *via* educational attainment is significantly different
from zero ($p < 0.05/33$).

@@

Previous work has documented natural selection in modern populations on variants
underlying polygenic traits [@beauchamp2016genetic; @kong2017selection;
@sanjak2018evidence]. We show that correlations between polygenic scores and
fertility are highly concentrated among specific subgroups of the population,
including people with lower income, lower education, younger first parenthood,
and more lifetime sexual partners. Among older mothers (22+), selection effects
are reversed. Furthermore, the size of selection effects on a polygenic score
correlates with that score's association with labour market earnings. The
economic theory of fertility provides a parsimonious explanation for these findings.
Because of the substitution effect of earnings on fertility, scores are selected
for when they correlate with low human capital, and this effect is stronger at 
lower levels of income and education.

@@

If inequalities in polygenic scores are important for understanding social
structure and mobility [@belsky2018genetic;
@Rimfeld_2018; @harden2021genetic], then these increases are substantive. A
back-of-the-envelope calculation, using estimated heritability for educational
attainment of 0.4 [@branigan2013variation], gives the proportion of variance in
income attributable to the "true" polygenic score for educational attainment as
`r r2_income_true_psea` among respondents. Natural selection would raise this
proportion to `r r2_income_true_psea_wt` among respondents' children,
counterfactually holding environmental conditions constant. Similarly, since
many polygenic scores are predictive of disease risk, they could potentially
increase health inequalities. In general, the evolutionary history of
anatomically modern humans is related to disease risk [@benton2021influence];
understanding the role of contemporary natural selection may aid in mapping the
genetic architecture of current health disparities.

>>>>

We created polygenic scores for 33 traits in `r comma_num(nrow(famhist))`
individuals of European descent, corrected for ancestry using 100 genetic
principal components (see Materials and Methods). Figure
\@ref(fig:plot-means-over-time) plots mean polygenic scores in the sample by
5-year birth intervals. Several scores show consistent increases or declines
over this 30-year period, of the order of 5% of a standard deviation. These
changes could reflect natural selection within the UK population, but also
emigration, or ascertainment bias within the sample [@10.1093/aje/kwx246].

@@

To show this, we write down a simple model of fertility choices. $h$ is
an individual's level of human capital. For now, we simply identify this with
his or her wage $W$. Raising a child takes time $b$. People maximize utility $U$
from the number of children $N$ and from income $Y\equiv(1-bN)W$:

@@

res_mediation %<>% mutate(
                     prop = estimate_ind/estimate_total,
                     # 0.025 not 0.05 because two-sided
                     sig  = abs(statistic_ind) > qnorm(1 - 0.025/33)
                   )

@@

median_prop <- median(res_mediation$prop[res_mediation$prop > 0 & 
                                           res_mediation$sig])

@@

Second, we run a mediation analysis to directly test whether the correlation
between each polygenic score and fertility is mediated by educational attainment
(Appendix Table \@ref(tab:tbl-res-mediation)). For `r n_pos_sig` scores, the
indirect effect of the score on fertility *via* educational attainment is
significantly different from zero ($p < 0.05/33$). Among these scores, the
median proportion of the total effect explained by the indirect effect is 
`r percent(median_prop)`.

@@

Previous work has documented natural selection in modern populations on variants
underlying polygenic traits [@beauchamp2016genetic; @kong2017selection;
@sanjak2018evidence]. We show that correlations between polygenic scores and
fertility are highly concentrated among specific subgroups of the population,
including people with lower income, lower education, younger first parenthood,
and more lifetime sexual partners. Among mothers aged 22+, selection effects
are reversed. Furthermore, the size of selection effects on a polygenic score
correlates with that score's association with labour market earnings. The
economic theory of fertility provides a parsimonious explanation for these findings.
Because of the substitution effect of earnings on fertility, scores are selected
for when they correlate with low human capital, and this effect is stronger at 
lower levels of income and education.

@@

If inequalities in polygenic scores are important for understanding social
structure and mobility [@belsky2018genetic;
@Rimfeld_2018; @harden2021genetic], then these increases are substantive. A
back-of-the-envelope calculation, using estimated heritability for educational
attainment of 0.4 [@branigan2013variation], gives the proportion of variance in
household income attributable to the "true" polygenic score for educational
attainment as `r r2_income_true_psea` among respondents. Natural selection would
raise this proportion to `r r2_income_true_psea_wt` among respondents' children,
counterfactually holding environmental conditions constant. Similarly, since
many polygenic scores are predictive of disease risk, they could potentially
increase health inequalities. In general, the evolutionary history of
anatomically modern humans is related to disease risk [@benton2021influence];
understanding the role of contemporary natural selection may aid in mapping the
genetic architecture of current health disparities.
